---
title: "Prosper Loan Data EDA"
author: "Shahrooz Govahi"
date: "March 12, 2017"
output: html_document
---

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(grid)
library(gridExtra)
library(lubridate)
library(GGally)
library(dplyr)
library(psych)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data

setwd("C:/Users/Shahrooz/Desktop/P4-R")
# putting null values as "NA"s in the data set.
loan_Orig <- read.csv("prosperLoanData.csv", na.strings = c("", "NA"))

# creating a new variable for CreditScore base on UpperRange and LowerRange 
# of CreditScore
loan_Orig$CreditScore <- (loan_Orig$CreditScoreRangeLower + 
                                loan_Orig$CreditScoreRangeUpper)/2 + 0.5

# calculating CreditAge

# converting dates to POSIX format for doing calculation with it
loan_Orig$DateCreditPulled <- as.POSIXlt(loan_Orig$DateCreditPulled)
loan_Orig$FirstRecordedCreditLine <- 
                          as.POSIXlt(loan_Orig$FirstRecordedCreditLine)
# adding CredutAge as a new variable to dataframe
loan_Orig$CreditAge <- year(loan_Orig$DateCreditPulled) - 
                                year(loan_Orig$FirstRecordedCreditLine)
```

Base on LoanStatus categories; I think one possible goal for this EDA should be
looking at different features and predict either that loan will be Completed or
Failed. By Failed loans I am referring to chargedoff or Defaulted loans. The 
other categories are not useful for this purpose. So I am defining a new 
variable LoanStatusLabel. I will put it's value as Completed if it's Completed 
and Failed if it's Defaulted or Chargedoff and NA for other categories
(e.g.Cancelled, Current).

```{r echo = FALSE, message=FALSE, warning=FALSE, LoanStatusLabel}
### creating a new variable base on LoanStatus
### Don't forget that if statement is not a vectorized function and that's
### the reason I wrote it in this way; Something like nested if statements.
loan_Orig$LoanStatusLabel <- with(loan_Orig, 
                                  ifelse(LoanStatus == "Chargedoff", "Failed",
                                  ifelse(LoanStatus == "Defaulted", "Failed",
                                  ifelse(LoanStatus == "Completed", "Completed", NA))))

### converting LoanStatusLabel as factor
loan_Orig$LoanStatusLabel <- as.factor(loan_Orig$LoanStatusLabel)
```

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Creating_SubSet}
# Dimention of the original dataset
# dim(loan)


# The choosen features out of 81 variables and the subset data structure we have
my_variables <- c("CreditGrade", "Term", "BorrowerAPR", "ProsperScore", 
                  "BorrowerState","EmploymentStatus", "IsBorrowerHomeowner", 
                  "CreditScore", "CreditAge", "OpenCreditLines", 
                  "CurrentDelinquencies", "BankcardUtilization", 
                  "DebtToIncomeRatio", "ProsperPrincipalBorrowed", 
                  "LoanOriginalAmount", "LoanStatusLabel")
loan <- loan_Orig[my_variables]
# Omitting rows with NA for LoanStatusLabel; Focusing on Completed of Failed 
# Loans only
loan <- subset(loan, !is.na(LoanStatusLabel))
## another way for subsetting
## loan <- subset(loan_Orig, select = c("OpenCreditLines"))
## removing loan_Orig dataframe
rm(loan_Orig)
rm(my_variables)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Credit_Grade}
# looking at different credit Grades and their count in the dataset
qplot(data = subset(loan, !is.na(CreditGrade)), x = CreditGrade,
      color = I('black'), fill = I('blue'))
summary(loan$CreditGrade)
```

The majority of Credit Grades in our data set are NAs; Most of Credit Grades 
are C and the least number of them are NC.

```{r echo=FALSE, message=FALSE, warning=FALSE, Term}
# Term has only 3 different numbers and it's better to look at it as factors 
# rather than integers
loan$Term <- as.factor(loan$Term)
qplot(x = Term, data = loan, color = I('black'), fill = I('yellow'))
summary(loan$Term)
```

The majority of loan terms are 36 months and less than 2% of them have 12 months 
term.

```{r echo = FALSE, message=FALSE, warning=FALSE, BorrowerAPR}
p1 = qplot(data = subset(loan, !is.na(BorrowerAPR)), x = BorrowerAPR,
          binwidth = 0.005,
          xlab = "Borrower APR",
          ylab = "Number of Borrowes",
          color = I('black'), fill = I('green')) +
      scale_x_continuous(limits = c(0.05, 0.45), breaks = seq(0.05, 0.45, 0.02))

p2 = qplot(data = subset(loan, !is.na(BorrowerAPR)), 
          x = log10(BorrowerAPR + 1),
          binwidth = 0.0015,
          xlab = "Log10 of Borrower APR",
          ylab = "Number of Borrowes",
          color = I('black'), fill = I('red'))

p3 = qplot(data = subset(loan, !is.na(BorrowerAPR)), x = sqrt(BorrowerAPR),
          binwidth = 0.01,
          xlab = "Sqrt of Borrower APR",
          ylab = "Number of Borrowes",
          color = I('black'), fill = I('blue'))

grid.arrange(p1, p2, p3, ncol = 1)

summary(loan$BorrowerAPR)
```

The most frequent Borrower APR is 0.36;
By transforming data with Square root, the distribution result is more smoother.

```{r echo=FALSE, message=FALSE, warning=FALSE, APRfreqpoly}
qplot(data = subset(loan, !is.na(BorrowerAPR)), 
      x = BorrowerAPR,
      geom = 'freqpoly')
```

```{r echo = FALSE, message=FALSE, warning=FALSE, ProsperScore}
qplot(data = subset(loan, !is.na(ProsperScore)), x = ProsperScore) +
  scale_x_continuous(limits = c(0, 12), breaks = 1:12)

summary(loan$ProsperScore)
```

The most frequesnt ProspectScores are 4, 6 and 8; On the other hand, 1 and 11
are the least frequent ones.

```{r echo = FALSE, message=FALSE, warning=FALSE, BorrowerState}
### learned from Reviwer:For changing the text size or text angle in labels, 
### use: theme()
qplot(data = subset(loan, !is.na(BorrowerState)), x = BorrowerState,
      fill = I('red')) +
  theme(text = element_text(size=8),
        axis.text.x = element_text(angle = 90))
  
summary(loan$BorrowerState)
```

California has the most users of Prosper loan services.

```{r echo = FALSE, message=FALSE, warning=FALSE, EmploymentStatus}
### learned from Reviwer:for making smaller values visible, I used log for y 
### axis;
qplot(data = subset(loan, !is.na(EmploymentStatus)), 
      x = EmploymentStatus, fill = I('purple'), ylab = 'log10 (Frequency)') +
  scale_y_log10()
summary(loan$EmploymentStatus)
```

Employed category is the most frequent employment status.

```{r echo = FALSE, message=FALSE, warning=FALSE, IsBorrowerHomeowner}
qplot(data = subset(loan, !is.na(IsBorrowerHomeowner)), x = IsBorrowerHomeowner,
      color = I('black'), fill = I('red'))
summary(loan$IsBorrowerHomeowner)
```

Home owners and not home owners numbers are very close to each other;

```{r echo = FALSE, message=FALSE, warning=FALSE, CreditScore}
qplot(data = subset(loan, !is.na(CreditScore)), x = CreditScore,
      color = I('white'), fill = I('orange'))
summary(loan$CreditScore)
```

The CreditScore = 10 should be an error in our data; So I will focus on scores
more than 460;

```{r echo = FALSE, message=FALSE, warning=FALSE, scaled_CreditScore}
qplot(data = subset(loan, !is.na(CreditScore)), x = CreditScore,
      color = I('white'), fill = I('orange')) +
  scale_x_continuous(limits = c(460, 900), breaks = seq(460, 900, 20))
```

There are some gaps between scores. e.g. There is no person with credit score 
equal to 700 or 560. It seems like credit scores has 7 clusters.

```{r echo=FALSE, message=FALSE, warning=FALSE, CreditScorefreqpoly}
qplot(data = subset(loan, !is.na(CreditScore)),
      x = CreditScore, y = ..count../sum(..count..),
      geom = 'freqpoly', color = 'red', xlim = c(425, 900))
```


```{r echo = FALSE, message=FALSE, warning=FALSE, CreditAge}
qplot(data = subset(loan, !is.na(CreditAge)), x = CreditAge, binwidth = 2,
      color = I('black'), fill = I('blue')) +
  scale_x_continuous(breaks = seq(0, 60, 2))
```

The CreditAge is a little right skewed. So I will look at Square root of the
data as well.

```{r echo = FALSE, message=FALSE, warning=FALSE, Sqrt_CreditAge}
qplot(data = subset(loan, !is.na(CreditAge)), x = sqrt(CreditAge),
      color = I('black'), fill = I('green'))
summary(loan$CreditAge)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, OpenCreditLine}
qplot(data = subset(loan, !is.na(OpenCreditLines)), x = OpenCreditLines,
      binwidth = 1,
      color = I('black'), fill = I('red')) +
  scale_x_continuous(breaks = seq(0, 40, 1))
```

The OpenCreditLines is a little skewed. I tried transforming it with log10
bot the result is not very useful.

```{r echo = FALSE, message=FALSE, warning=FALSE, log10_OpenCreditLine}
qplot(data = subset(loan, !is.na(OpenCreditLines)), 
      x = log10(OpenCreditLines + 1),
      color = I('black'), fill = I('red'))
summary(loan$OpenCreditLines)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, CurrentDelinquencies}
qplot(data = subset(loan, !is.na(CurrentDelinquencies)), 
      x = CurrentDelinquencies,
      color = I('black'), fill = I('purple')) +
  scale_x_continuous(breaks = seq(0, 20, 2))
summary(loan$CurrentDelinquencies)
```

Most of the users has no Current Delinquency.

```{r echo = FALSE, message=FALSE, warning=FALSE, BankcardUtilization}
qplot(data = subset(loan, !is.na(BankcardUtilization)), x = BankcardUtilization,
      binwidth = 0.05) +
  scale_x_continuous(limits = c(-0.1, 1.5), breaks = seq(-0.1, 1.5, 0.1))
summary(loan$BankcardUtilization)
```

BankcardUtilization has a bimodal distribution.
Most users utilize 0 percent of their credit cards and next the majority of 
users utilize 95% of their bank cards.

```{r echo = FALSE, message=FALSE, warning=FALSE, DebtToIncomeRatio}
qplot(data = subset(loan, !is.na(DebtToIncomeRatio)), 
      x = DebtToIncomeRatio,
      color = I('black'), fill = I('green')) +
  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 0.25))

nrow(subset(loan, DebtToIncomeRatio > 5))
```

I think the DebtToIncomeRatio greater than 10 is a problem in dataset. Because
there are totally 270 points with this ratio greater 6 and 247 of them are 
greater than 10;
There are a few outliers in DebtToIncomeRatio, so I will change the scales to 
avoiding them;

```{r echo = FALSE, message=FALSE, warning=FALSE, scaled_DebtToIncomeRatio}
qplot(data = subset(loan, !is.na(DebtToIncomeRatio)), 
      x = DebtToIncomeRatio,
      binwidth = 0.1,
      color = I('black'), fill = I('green')) +
  scale_x_continuous(limits = c(0, 1.5), breaks = seq(0, 1.5, 0.1))
summary(loan$DebtToIncomeRatio)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, ProsperPrincipalBorrowed}
qplot(data = subset(loan, !is.na(ProsperPrincipalBorrowed)), 
      x = ProsperPrincipalBorrowed,
      color = I('black'), fill = I('blue')) +
  scale_x_continuous(limits = c(0, 45000), breaks = seq(0, 45000, 2000))
```

Most of the loans are under $6000. The distribution is right skewed;

```{r}
loan_under_6000 <- subset(loan, ProsperPrincipalBorrowed <= 6000)
nrow(loan_under_6000)
```


```{r echo = FALSE, message=FALSE, warning=FALSE, log10_ProsperPrincipalBorrowed}
qplot(data = subset(loan, !is.na(ProsperPrincipalBorrowed)), 
      x = log10(ProsperPrincipalBorrowed + 1), binwidth = 0.1,
      color = I('black'), fill = I('blue'), xlim = c(3, 5))

summary(loan$ProsperPrincipalBorrowed)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, LoanOriginalAmount}
qplot(data = subset(loan, !is.na(LoanOriginalAmount)), 
      x = LoanOriginalAmount,
      binwidth = 1000,
      color = I('black'), fill = I('blue'))
```

There are some ouliers in LoanOriginalAmount, so I changed the scales.

```{r echo = FALSE, message=FALSE, warning=FALSE, scaled_LoanOriginalAmount}
qplot(data = subset(loan, !is.na(LoanOriginalAmount)), 
      x = LoanOriginalAmount,
      binwidth = 1000,
      color = I('black'), fill = I('blue')) +
  scale_x_continuous(limits = c(1000, 16000), breaks = seq(1000, 16000, 1000))

summary(loan$LoanOriginalAmount)
```

The most popular LoanOriginalAmounts are between $2500 and $3500.

```{r echo = FALSE, message=FALSE, warning=FALSE, LoanStatLabel}

qplot(data = subset(loan, !is.na(LoanStatusLabel)), x = LoanStatusLabel,
      color = I('black'), fill = I('green'), ylim = c(0, 60000))

table(loan$LoanStatusLabel)
```


# Univariate Analysis

### What is the structure of your dataset?

The original data set contains 113,937 observations with 81 variables. 
I choose some of those variablels and focusing only on Completed or failed loans
for doing EDA;
It means that I am working on 55084 observations with 16 variables.

```{r echo = FALSE, message=FALSE, warning=FALSE, final_dataset_Structure}
dim(loan)
str(loan)
summary(loan)
```


### What is the main feature of interest in your dataset?
In this investigation I want to consider final result of a loan: 
LoanStatusLabel as the main feature of interest and I am looking at other 
features as predictors of this target.

### What other features in the dataset do you think will help support your
### investigation into your feature(s) of interest?
I choose these 15 features to help my investigation:
CreditGrade, Term, ProsperScore, BorrowerAPR, BorrowerState, EmploymentStatus,
IsBorrowerHomeowner, CreditScore, CreditAge, OpenCreditLine, 
CurrentDelinquencies, BankcardUtilization, DebtToIncomeRatio, 
ProsperPrincipalBorrowed, LoanOriginalAmount. 

### Did you create any new variables from existing variables in the dataset?
I created these variables:
1. CreditAge, base on FirstRecordedCreditLine and DateCreditPulled
2. CreditScore, base on CreditScoreRangeLower and CreditScoreRangeUpper I 
calculated the average of these two ranges to use one variable. 
3. LoanStatusLabel, base on LoanStatus I create 2 categories:
   Completed, Failed.

### Of the features you investigated, were there any unusual distributions?
The EmploymentStatus is right skewed. The reason should be that, the prosper
company only focus on Employed or Full_time or Self-employed applicants. 
Therefore the majority of it's clients are in these three categories and very 
few of them are in retired or other kinds of employment types.
BankcardUtilization has a bimodal distribution. One peak occurs at 0 utilization
and the other is at 95% utilization.

### Did you perform any operations on the data to tidy, adjust, or change the
### form of the data? If so, why did you do this?
*I only look at two kinds of LoanStaus, because I am trying to find out what's
the most effective features for predicting a Completed loan. So I cleaned all
other LoanStatuses.
* I changed the Term variable as factor. Because it has only 3 fixed values:
12, 36, 60.
*For all graphs, I put NA's aside for looking at values only.
* Another issue was distribution type for some of the features, like 
BorrowerAPR. It does not have a normal distribution. So I tried log10 and sqrt 
of this variable, and I think sqrt of values is smoother and more similar to 
normal distribution.

# Bivariate Plots Section

### Correlations

```{r echo=FALSE, message=FALSE, warning=FALSE, CorrelationCoef}
loan_for_cor <- subset(loan, select = c(BorrowerAPR, ProsperScore, CreditScore,
                                        CreditAge, OpenCreditLines,
                                        CurrentDelinquencies, 
                                        BankcardUtilization, DebtToIncomeRatio,
                                        ProsperPrincipalBorrowed, 
                                        LoanOriginalAmount))
#cor(loan_for_cor, method = "pearson", use = "complete.obs")
ggcorr(loan_for_cor, nbreaks = 5)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, scatterplotMatrix}
set.seed(42)
loan_for_graph <- subset(loan, select = c(BorrowerAPR, ProsperScore, CreditScore,
                                          CreditAge, CurrentDelinquencies, 
                                          BankcardUtilization, DebtToIncomeRatio,
                                          ProsperPrincipalBorrowed, 
                                          EmploymentStatus))
# ggpairs(loan_for_graph[sample.int(nrow(loan_for_graph), 1000), ])
# ggsave('manyplots.jpg')
# Better and cleaner way
pairs.panels(loan_for_graph[sample.int(nrow(loan_for_graph), 1000), ], 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
```


```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusLabelBorrowerAPR}
qplot(data = loan,
      x = LoanStatusLabel, y = BorrowerAPR,
      geom = 'boxplot')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, loanStatusSummary}
by(loan$BorrowerAPR, loan$LoanStatusLabel, summary)
```

The APR mean for failed loans is 4.5% more than completed loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, CurrentDelinquenciesBox}
qplot(x = CurrentDelinquencies, data = loan, y = ..count../sum(..count..),
  geom = 'freqpoly', color = LoanStatusLabel)
```

Around 60% of completed loans has zero CurrentDelinquencies and this number is
20% for failed loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusLabelDelinquencies}
qplot(data = loan,
      x = LoanStatusLabel , y = CurrentDelinquencies, ylim = c(0, 5),
      geom = 'boxplot')
```


```{r echo=FALSE, message=FALSE, warning=FALSE, DelinquenciesSummary}
by(loan$CurrentDelinquencies, loan$LoanStatusLabel, summary)
```

Mean of CurrentDelinquencies for Colpleted loans is 0.6 and this number is 1.6
for failed loans.
CurrentDelinquencies is a powerful predictor for failing a loan.

```{r echo=FALSE, message=FALSE, warning=FALSE, DebtToIncomeRatiofreq}
qplot(x = DebtToIncomeRatio, data = loan, y = ..count../sum(..count..),
  geom = 'freqpoly', color = LoanStatusLabel) +
  scale_x_continuous(limits = c(0, 2.5), breaks = seq(0, 2, 0.2))
```

DebtToIncomeRatio is less than 20% for 24% of completed loans and this number is
around 8% for failed loans. 

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusLabelDebtToIncome}
qplot(data = loan,
      x = LoanStatusLabel , y = DebtToIncomeRatio, ylim = c(0, 5),
      geom = 'boxplot', fill = I("#ff1a66"))

by(loan$DebtToIncomeRatio, loan$LoanStatusLabel, summary)
```

So DebtToIncomeRatio is another good predictor for loan status.

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusLabelPlot}
qplot(data = loan,
      x = BorrowerAPR, y = ..count../sum(..count..), xlim = c(0.4, 0.6),
      geom = 'freqpoly', color = LoanStatusLabel)

qplot(data = loan,
      x = CreditScore, y = ..count../sum(..count..),xlim = c(460, 890),
      geom = 'freqpoly', color = LoanStatusLabel)

qplot(data = loan,
      x = CreditAge, y = ..count../sum(..count..),
      geom = 'freqpoly', color = LoanStatusLabel)

loan_high_APR <- subset(loan, BorrowerAPR > 0.42)
table(loan_high_APR$LoanStatusLabel)
```

It seems like most of loans with APR more than 0.45 failed.

```{r echo=FALSE, message=FALSE, warning=FALSE, loanStatusAge}
loan_status_Age <- loan %>%
  group_by(LoanStatusLabel) %>%
  summarise(CreditAge_mean = mean(CreditAge, na.rm = TRUE),
            CreditAge_median = median(CreditAge, na.rm = TRUE),
            n = n())
loan_status_Age
```

The difference between CreditAge mean and median for 2 group of LoanStatusLabel
of interest (Completed and Failed) is not very significant.

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusScore}
loan_status_Score <- loan %>%
  group_by(LoanStatusLabel) %>%
  summarise(creditScore_mean = mean(CreditScore, na.rm = TRUE),
            CreditScore_median = median(CreditScore, na.rm = TRUE),
            n = n())
loan_status_Score
```

The median CreditScore for Completed and Failed LoanStatusLables has 40 points 
difference.

```{r echo=FALSE, message=FALSE, warning=FALSE, BorrowerAPRWrapStatus}
qplot(data = subset(loan, !is.na(BorrowerAPR)), x = BorrowerAPR,
      color = I('black'), fill = I('green'))+
  facet_wrap(~ LoanStatusLabel, ncol = 1)
```

The distribution of different result for LoanStatusLabel in BorrowerAPR shows 
that Completed loans are a little right skewed and Faild loans are roughly 
uniformly distributed. It means most of the failed loans relativly had larger 
APRs.

```{r echo=FALSE, message=FALSE, warning=FALSE, APRProsperScore}
### Learened from Reviewer: When on variable is factor scatterplot is not
### appropriate; use boxplot
ggplot(aes(y = BorrowerAPR, x =factor(ProsperScore)), 
       data = subset(loan, !is.na(ProsperScore))) +
  geom_boxplot(alpha = 1/20)
```

The ProsperScore and BorrowerAPR are negatively correlated 
with -0.74 Correlation Coefficient. That's a strong correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, corAPRProsperScore}
cor(loan$BorrowerAPR, loan$ProsperScore, use = 'complete')
```


```{r echo=FALSE, message=FALSE, warning=FALSE, CreditGradeCreditScore}
ggplot(aes(x = CreditGrade, y = CreditScore), 
       data = subset(loan, !is.na(CreditGrade) & CreditGrade != 'NC' & 
                       CreditScore > 400),
       ylim = c(460, 890)) +
  geom_boxplot(alpha = 1/20, fill = I('#ff0000'))
```

There are many overlaps, so I am using Jitter to see more clearly.

```{r echo=FALSE, message=FALSE, warning=FALSE, jitterPlot}
ggplot(aes(CreditGrade, CreditScore), 
       data = subset(loan, !is.na(CreditGrade))) +
  geom_jitter(alpha = 1/30, color = "#4EFAE3")
```

It seems like CreditGrades are implemented base on CreditScore.

```{r echo=FALSE, message=FALSE, warning=FALSE, APRCreditScore}
### Learned from Reviewer: Put a regression line for showing the trend.
ggplot(aes(x = BorrowerAPR, y = CreditScore), data = loan) +
  geom_point(alpha = 1/35, color = "#8080ff") +
  geom_smooth(method = 'lm', color = 'red')
```

There is a negative correlation between BorrowerAPR and CreditScore:

```{r echo=FALSE, message=FALSE, warning=FALSE, corAPRCreditScore}
### For omitting NAs I put use="complete";
cor(loan$BorrowerAPR, loan$CreditScore, use="complete")
```


```{r echo=FALSE, message=FALSE, warning=FALSE, CreditScoreProsperScorePlot}
ggplot(aes(y = CreditScore, x = ProsperScore), data = loan) +
  geom_jitter(alpha = 1/20, color = I('green')) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_y_continuous(limits = c(520, 800), breaks = seq(520, 800, 50))
```

ProsperScore and CreditScore variables are positively correlated;
The correlation coefficient is:

```{r echo=FALSE, message=FALSE, warning=FALSE, corCreditScoreProsperScore}
cor(loan$CreditScore, loan$ProsperScore, use = 'complete')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, CreditScoreProsperScoreHeatMap}
### HeatMap
ggplot(aes(y = CreditScore, x = ProsperScore, fill = CreditAge),
       data = loan) +
       geom_tile() +
       scale_fill_gradientn(colors = colorRampPalette(c('blue', 'red'))(100))
```


Term and LoanOriginalAmount:

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanAMountTerm}
ggplot(aes(x = Term, y = LoanOriginalAmount), data = loan) +
  geom_jitter(alpha = 1/20, color = I("#2929a3"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, APRDebtToIncomePlot}
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerAPR), data = loan) +
  geom_point(alpha = 1/10, color = I("#e60099"))
```

Correlation between DebtToIncomeRatio and BorrowerAPR:

```{r echo=FALSE, message=FALSE, warning=FALSE, corAPRDebtToIncome}
cor(loan$DebtToIncomeRatio, loan$BorrowerAPR, use = 'complete')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, CreditAgeCreditScorePlot}
### Learned from Reviewer
ggplot(aes(x = CreditAge, y = CreditScore), data = loan) +
  geom_point(alpha = 1/40, color = I('#0033cc')) +
  geom_smooth(method = 'lm', color = 'red') +
  ylim(300, 850)
```

Correlation between CreditAge and CreditScore:

```{r echo=FALSE, message=FALSE, warning=FALSE, corCreditAgeCreditScore}
cor(loan$CreditAge, loan$CreditScore, use = 'complete')
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. How did the feature of interest vary with other features
### in the dataset?
* Lower BorrowerAPR has a bigger change of being a Completed loan rather than
the larger BorrowerAPRs. The majority of completed loans have BorrowerAPR less
than 0.2
* Borrowers with more CurrentDelinquencies are more likely to fail a loan rather
than borrowers with zero number of CurrentDelinquencies.
* Borrower that complete their loan, has a bigger proportion of larger 
CreditScore rather than failed loan borrowers.

### Did you observe any interesting relationships between the other features
### (not the main feature of interest)?
There are some strong relationships between CreditScore and many other variables
like: BankcardUtilization, DebtToIncomeRatio, CurrentDelinquencies. It makes
sense. Because CreditScore is a function of many of these variables. On the 
other hand ProsperScore has good correlation with CreditScore, and it should be
because they both have same features for creation.

### What was the strongest relationship you found?
The strongest correlation is between ProsperScore and BorrowerAPR with 
correlation coefficient equal to -0.73; That's maybe related to this fact that 
one of the main patameters for determining BorrowerAPR is ProsperScore;

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = BorrowerAPR), data = loan) +
  geom_histogram(aes(fill = LoanStatusLabel)) +
  facet_wrap(~ CreditScore)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, BorrowerAPRCreditScoreBoxPlot}
### Learned from Reviewer
ggplot(aes(x = factor(CreditScore), y = BorrowerAPR), 
       data = subset(loan, !is.na(CreditScore) & CreditScore > 400)) +
  geom_boxplot(alpha = 1/20, fill = I('yellow'))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, loan_Good_Bad_Credit}
loan_good_credit <- subset(loan, CreditScore >= 810)
table(loan_good_credit$LoanStatusLabel)

loan_bad_credit <- subset(loan, CreditScore <= 510)
table(loan_bad_credit$LoanStatusLabel)
```

Above 90% of debtors with creditScore more than 810, are Completed their Loan 
and the majority(69%) of debtors with CreditScore under 510, failed their loan.

```{r echo = FALSE, message=FALSE, warning=FALSE, DebtToIncome_ProsperScore}
loan_good_area <- subset(loan, DebtToIncomeRatio < 0.5 & ProsperScore > 8)
table(loan_good_area$LoanStatusLabel)
```

Loans with DebtToIncomeRatio less than 0.5 with ProsperScore more than 8 is very 
likely to complete(91%).

```{r echo = FALSE, message=FALSE, warning=FALSE, DebtToIncomeProsperScoreBoxPlot}
### Learned from Reviewer
loan$LoanOriginalAmount.bucket = cut(loan$LoanOriginalAmount,
                            c(0, 10000, 20000, 30000))

ggplot(aes(y = (DebtToIncomeRatio), x = factor(ProsperScore)), 
       data = subset(loan, !is.na(LoanOriginalAmount.bucket) & 
                       !is.na(ProsperScore)))+
  geom_boxplot(aes(fill =  LoanOriginalAmount.bucket)) + ylim(0, 1)
```

When DebtToIncomeRatio decreases, the ProsperScore increases; For ProsperScore
less than 3, there is no LoanOriginalAmount greater than 20000.

```{r echo = FALSE, message=FALSE, warning=FALSE, BorrowerAPR_ProsperScore}
ggplot(aes(x = BorrowerAPR, y= ProsperScore), data = loan) +
  geom_jitter(aes(color = LoanStatusLabel, shape = EmploymentStatus)) +
  facet_wrap( ~ Term)

loan_low_APR <- subset(loan, BorrowerAPR < 0.2 & Term == 12 & 
                         (EmploymentStatus == "Employed" | 
                            EmploymentStatus == "Full-time"))

table(loan_low_APR$LoanStatusLabel)
```

The most successful loans has 12 months Term with BorrowerAPR less than 0.2 and 
Employed or Fulltime borrower. 98% of these loans are Completed.

```{r echo = FALSE, message=FALSE, warning=FALSE, EmploymentStatus_OpenCreditLines}
ggplot(aes(x = EmploymentStatus , y = OpenCreditLines), data = loan) +
  geom_jitter(aes(color = LoanStatusLabel)) +
  facet_wrap( ~ IsBorrowerHomeowner)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, EmploymentStatus_CreditScore}
ggplot(aes(x = EmploymentStatus , y = CreditScore), data = loan) +
  scale_y_continuous(limits = c(400, 900)) +
  geom_jitter(aes(color = LoanStatusLabel)) +
  facet_wrap( ~ IsBorrowerHomeowner)
```

HomeOwners has larger margings for CreditSCore and if they are Employed or 
Fulltime job, it's more likely to complete their loan.

```{r echo = FALSE, message=FALSE, warning=FALSE, BorrowerAPR_LonaAmount}
ggplot(aes(x = BorrowerAPR , y = LoanOriginalAmount), data = loan) +
  geom_line(aes(color = LoanStatusLabel), stat = 'summary', fun.y = median) +
  geom_smooth() +
  facet_wrap( ~ IsBorrowerHomeowner)
loan_home <- subset(loan, LoanStatusLabel == "Failed" & 
                      LoanOriginalAmount > 10000)
table(loan_home$IsBorrowerHomeowner)
```

In general HomeOwners are qualified for larger loans and the percentage of their
failure is more than not HomeOwners.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. Were there features that strengthened each other in terms of
### looking at your feature of interest?
Terms of loans and BorrowerAPR and EmploymentStatus are creat a good predictor
for LoanStatusLabel. e.g. a loan with 12 months Term and APR less than 0.2 for
a borrower that is Employed or has Fulltime job is very likely to complete.

### Were there any interesting or surprising interactions between features?
HomeOwners with Original Loan amount more than 10000 are more likely to fail 
thier prosper loan rather than not HomeOwners the ratio is 67%.

### OPTIONAL: Did you create any models with your dataset? Discuss the 
### strengths and limitations of your model.
I think with some of these features we can train a classifier for predicting
Completed of Failed loans.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
qplot(data = subset(loan, !is.na(BankcardUtilization)), x = BankcardUtilization,
      binwidth = 0.05,
      xlab = 'Bank Card Utilization[%]',
      ylab = 'Number of Members',
      main = 'Bank Card Utilization Distribution',
      color = I('black'), fill = I('yellow')) +
  scale_x_continuous(limits = c(-0.1, 1.5), breaks = seq(-0.1, 1.5, 0.1))
```

### Description One

It's a bimodal distribution. Many of members have 0 revolving bank card 
utilization and the second crowded point is 0.95 utilization. It's interesting
because there are two peaks in two sides of the spectrum.
The first quartile is 0.21 and mean of this variable is 0.53. The maximun value
of this variable is 5.95;

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
qplot(data = loan,
      x = LoanStatusLabel, y = BorrowerAPR,
      xlab = 'Loan Status',
      ylab = 'Borrower APR[%]',
      main = 'Borrower APR Vs. Different Loan Statuses',
      geom = 'boxplot', fill = I("#00b359"))

#qplot(data = loan,
#      x = BorrowerAPR, y = ..count../sum(..count..), xlim = c(0.4, 0.6),
#      xlab = 'Borrower APR',
#      ylab = 'Proportion of Memberswith that APR',
#      main = 'APR more than 0.40 with related Proportion of Borrowers',
#      geom = 'freqpoly', color = LoanStatusLabel)

#qplot(data = subset(loan, !is.na(BorrowerAPR)), x = BorrowerAPR,
#      color = I('black'), fill = I('green'),
#      xlab = 'BorrowerAPR',
#      ylab = 'Number of Borrowers',
#      main = 'Distribution of Borrower APRs for Different Loan Statuses') +
#  facet_wrap(~ LoanStatusLabel, ncol = 1)

loan_low_APR_2 <- subset(loan, BorrowerAPR < 0.2)
table(loan_low_APR_2$LoanStatusLabel)
```

### Description Two
The BorrowerAPR is a strong predictor for the loan status. Mean of APR for
Completed loans is roughly 0.05 less than mean of APR for Failed loans. The 
majority of loans with the APR more than 0.42 are failed: 6 faild in total 7.
The distribution of APR for Completed and Failed loans are very different. The
Completed loans have a right skewed distribution and the Failed loans have a 
left skewed distribution. It means most of Completed loans have lower APRs and
most of the Failed loans have larger APRs.
80% of loans with APR less than 0.2 are completed.
Minimun APR in this dataset is 0.0063 and maximum value is 0.512; The average of
APR is 0.222;

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# ggplot(aes(x = BorrowerAPR, y= ProsperScore), data = loan) +
#  geom_jitter(aes(color = LoanStatusLabel, shape = EmploymentStatus)) +
#  facet_wrap( ~ Term) +
#  xlab('Borrower APR') +
#  ylab('Prosper Score') +
#  ggtitle('Borrower APR Vs. Prosper Score for Different Terms and
#           Different Employment Status with Loan Status')

### Learned from Reviewer
ggplot(aes(y = BorrowerAPR, x= factor(ProsperScore)), 
       data = subset(loan, !is.na(ProsperScore))) +
  xlab('Prosper Score(Number)') +
  ylab('Borrower APR[%]') +
  ggtitle('Borrower APR Vs. Prosper Score') +
  geom_boxplot(aes(fill = Term))

loan_low_APR <- subset(loan, BorrowerAPR < 0.2 & Term == 12 & 
                         (EmploymentStatus == "Employed" | 
                            EmploymentStatus == "Full-time"))

table(loan_low_APR$LoanStatusLabel)
```

### Description Three
The most successful loans have 12 months Term. On the other hand 80% of loans
with Borrower APR less than 0.2 are completed. Another important parameter is
Employment status. Best categories for Employment Status are Employed and 
Full-time. If we add up all these conditions together the loans will
have 98% chance of being completed. That's a really high chance.
There are 16491 borrower that are Employed and 24957 borrower has Full-time job;
Very few of borrowers are not employed: 561; It makes sense because when Prosper
choose somebody to offer a loan, they prefer that person has some kind of job;
------

# Reflection

The Prosper loan data set contains 113,937 clients with 81 variables. I choose
16 variables and focus on specific categories and I found 55084 clients in the
new dataset.
Some variables are related to scoring clients. Variables like CreditScore,
ProsperScore, CreditGrade. These scores has close relationship together.

I was struggling how I choose a handful of variables between 81  
variables from the original dataset. I choose the most independent and not IDs
and gradually decrease them till I reach 16 of them. Another problem was the
dataset has null values for NAs and it takes a while to understand why I can not
filterout NAs with !is.na(variable). Another thing that I made decision about
was LoanStatus variable. It contain many different categories and after
considering them, I understand I only need two labels for loans: Completed or
Failed and the other categories are not finished and I can not use them as 
label.

I believe finding main features for predicting the loan status is a great 
success. And my graphs show how strong these features predict the loan status.
This is important for implementing the classifier for future research.
Defining CreditAge and CreditScore as two new variables are another success of
this study. Finding the importance of Borrower APR from different aspects, is
another success that I personally like it.

The process I went through has these stages:
First I investigate each variable distribution and used some transformations for
changing some non-normal distributions.
Next I investigate the relationship and correlations between different features.
Gradually I defined the project goal, finding the most important predictors for
successfull loans or Completed ones. I realize that The BorrowerAPR and Terms of
loan and ProsperScore and Employment Status are good predictors for categorizing
loans in Completed and Failed groups.

The next step base on these findings could be training a decision tree or naive
bayes classifier base on mentioned fetures for Loas Status as target.

Some limitation of this proccess should be the size of the filtered dataset. 
Because I only focus on Completed and Failed loan statuses and I can not more 
than 50% of the original dataset because those loans results are not finished 
and they are in progress. If I have access to the updated dataset, I should use 
all original datapoints for training and testing purposes and that really help 
when I double size of dataset.
